---
title: "Introduction to the mondate Package"
author: "Dan Murphy"
date: "November 23, 2015"
output: 
  pdf_document:
    fig_caption: yes
---

## Abstract:

Dates and the results of date arithmetic are 
exchanged throughout the world using the international standards 
"Data elements and interchange formats — Information interchange — Representation of dates and times" (ISO 8601:2004)
assembled by the International Organization for Standardization (ISO).
When accountants, actuaries and other financial services analysts
use the R Statistical Environment ("R") for creating models,
they have access to tried-and-true implementations of those standards
through R's offering of two "objects" ("classes") to the user:
POSIXt and Date.
Those two objects satisfy the vast majority of date/time user requirements.
After all, without them how could R have experienced its remarkable growth
in popularity
as the definitive go-to computational tool?

Those base objects are not completely satisfactory, however.
There are two key accounting concepts that 
are somewhat cumbersome to implement using base R alone:
*accounting date* and *account aging in month units*.

The purpose of this paper is show a way to
implement ISO-8601:2004-consistent solutions to those requirements in an R package.

## A Short Background on ISO 8601

It will be important to visit a few "basic concepts"
that most people understand intuitively but must be defined precisely
if computers are to understand too.

* time interval
* duration
* other?

### time interval

> Definition: time interval  
A time interval is "part of the time axis limited by two 
instants.
[It] comprises all instants between the two limiting instants 
and, unless otherwise stated, the limiting instants themselves."^[ISO 8601 2.1.3]

Figure 1 below is a diagram of two time intervals A and B on the time axis.
Each includes its respective endpoints
as indicated by the square brackets "[" and "]".
A and B overlap at time point 1.

```{r time_interval_int, fig.width = 10, fig.height=2, echo = F, fig.cap="a time interval"}
#plot.window(xlim=c(0,5), ylim=c(-.5, .5))
par(mar=c(1,2,1,2))
par(lend="butt")
plot(c(2,3), c(0,0), col="blue", type = "l", 
     axes=F, xlim=c(0,5), ylim=c(-.5, .5), 
     xlab="", ylab="")
axis(1, pos=0)
arrows(x0=-.1, y0=0, x1=5.15, length=.1)
text(5.1, 0, "t", pos = 1)
box()

segments(x0=0.01, y0=0, x1=0.99, col="blue", lwd =4)
text(x=0.013, y=0, "[", col="blue", font = 2, cex=2)
text(x=0.987, y=0, "]", col="blue", font = 2, cex=2)
text(x=0.5, y= .1, "A", col="blue")

segments(x0=1.01, y0=0, x1=1.99, col="blue", lwd =4)
text(x=1.013, y=0, "[", col="blue", font = 2, cex=2)
text(x=1.987, y=0, "]", col="blue", font = 2, cex=2)
text(x=1.5, y= .1, "B", col="blue")

#segments(x0=4.01, y0=0, x1=4.99, col="blue", lwd =4)
#text(x=4.01, y=0, "[", col="blue", font = 2, cex=2)
#text(x=4.987, y=0, "]", col="blue", font = 2, cex=2)
#text(x=4.5, y= .1, "C", col="blue")
```



Although allowed, 
it is generally undesirable for intervals to overlap
because of the ambiguity implied when a time instant 
is a member of two intervals.
This ambiguity arises because,
conforming with common tradition, 
all instants within an interval 
may be labeled by the same
"mark."^[See ISO 8601 2.1.4 Note 1.]

For example, suppose Figure 1 is actually a simple diagram of
the first days of calendar year 1970.
Intervals A and B correspond to January 1st and 2nd, respectively.
Indeed,
A is a "time interval" because it is a part of the time axis,
specifically the part between the two instants 
"midnight before January 1" and "midnight before January 2".
Any instant within interval A may wear the label "January 1, 1970"
equally in situations when the time of day is not important.
The ISO committee recognized the sharing of a common label among
similar neighboring objects at a more granular level when it said
"On a time scale consisting of successive steps [e.g., seconds], 
two distinct instants may be expressed by the same time 
point [e.g., 1970-01-01]."^[2.1.5 Note 1]

Under its broadest interpretation, however,
the sharing of labels among time intervals can create ambiguities,
particularly in the broadest case when
time intervals are allowed to be closed at both ends.

For example, 
intervals A and B overlap at midnight.
Does that instant belong to A or to B?
If "both", then wouldn't it follow that the midnight mark
"1970-01-02 00:00:00" could refer to the instants in both A and B?
What about when midnight is expressed as "1970-01-01 24:00:00"?^Also allowed by ISO 8601:2004

These complications can be easily avoided, however.
Simply leave 
one of the endpoints "open."

That is exactly what R does in its implementation of the 
`POSIXt` and `Date` classes.

Figure 2 is a plot of R's representation of the five days 
surrounding the transition from calendar year 1969 to 1970,
using the `POSIXct` class to express the endpoints.

```{r}
n = 5
a <- as.POSIXct((-n:n)*60*60*24, origin = "1970-01-01 00:00:00", tz="UTC")
a
```

```{r surrounding_5_days, fig.width = 10, fig.height=2, echo = F, fig.cap="beginning days of 1970, borders modeled by POSIXt instances"}
#plot.window(xlim=c(0,5), ylim=c(-.5, .5))
par(mar=c(1,2,1,2))
par(lend="butt")
NDaysInSecs <- 60*60*24*n
RadiusOfWindow <- NDaysInSecs + 60*60*24
plot(a, y=rep(0, 2*n+1), type = "l", axes = F, 
     ylim=c(-1, 1), xlim = c(-RadiusOfWindow, RadiusOfWindow),
     xlab="", ylab="")
axis(1, at=a, labels=a, pos=0)
arrows(x0=a[1]-60*60*24, y0=0, x1=a[2*n+1]+60*60*24, length=.1, code = 3)
text(a[2*n+1]+60*60*12, 0, "t", pos = 1)
segments(x0=a[1:(2*n)], y0=0, x1=a[2:(2*n+1)], col="blue", lwd =4)
for (i in 1:(2*n)) text(x=a[i] + 15*60, y=0, "[", col="blue", font = 2, cex=2)
for (i in 2:(2*n+1)) text(x=a[i] - 15*60, y=0, ")", col="blue", font = 2, cex=2)
axis(2, pos=0, labels = FALSE, lwd.ticks=0, lty="dotted")
text(0, .8, "origin =\n\"1970-01-01 00:00:00\"", pos=1)
box()
```

Labeling an interval

rotate around origin -- "dual"

"cut"

leap second?



### duration

> A duration is the "non-negative quantity ['magnitude'] attributed to a time interval, 
the value of which is equal to the difference between the time points 
of the final instant and the initial instant of the time interval."^[ISO 8601 2.1.6]

Think of a duration as the length of the vector ("magnitude") representing
a time interval.
In the figure below, the magnitude of the vector is one -- 
second, day, month, etc. --
depending on the context.

```{r duration, fig.width = 10, fig.height=2}
par(mar=c(1,2,1,2))
plot(2, 0, xlim=c(0,5), ylim=c(-.5, .5), col="green", pch = 20, axes=F, 
     xlab="", ylab="", main="duration")
arrows(x0=2, y0=0, x1=3, col="green", lwd =2)
box()
```

Under the International System of Units (SI),
the base unit of duration is "seconds".
Indeed,
whenever a duration is expressed in units other than "seconds"
that length of time is referred to as a "nominal duration."^[ISO 8601 2.1.7]
Examples of nominal durations are calendar day (its length in seconds can vary
depending on leap seconds and daylight/standard time shifts),
calendar month (its length in seconds varies, in addition to calendar day
variability, due to differing numbers of calendar days in a month),
and calendar year (in addition to calendar day variability, its length
can vary due to the addition of a leap day).
We will see examples of this behavior in R below.

#### seconds

`seconds` are the base unit for expressing duration.



As pertains
R,
that means ages need not rely on classes as granular as POSIXt;
the `Date` class, which keeps track of calendar days,
should suffice.

When a contract specifies an inception date but no time,
e.g., an insurance policy,
for 
the event E corresponding to the financial responsibilities incepting therein,
it is customary to consider those responsibilities to start at hour 00.
When considering the age of the aggregation of events occurring during a 
time interval, 
financial and accounting models will "tag" those events with the same 
instant of inception.
Two general instants can be found,
particularly in the actuarial literature:

1. the beginning instant of the interval
2. the midpoint of the interval.



It would seem that the calculation of an age in units of "days" would
be straightforward,
but there are a couple of complicating factors.
First, ISO 8601 defines a "day" to be the unit of time
equal to exactly 24 hours.
It defines a "calendar day" to be the interval of time
between successive midnights.
Those calendar day intervals are usually one day in length, 
but for two exceptions:

1. Leap seconds:
Every once in a while a second must be added to or subtracted
from Coordinated Universal Time (UTC) to realign UTC solar time (UT1).

2. Daylight time:
Frequently the clock is adjusted by local authorities.

`r 60*60*24 == 86400`
`r 86400*30 == 2592000`

`r as.Date("2015-12-01") - as.Date("2015-11-01") # Time difference of 30 days`

`r as.POSIXct("2015-12-01") - as.POSIXct("2015-11-01") # Time difference of 30.04167 days`

`r as.POSIXct("2015-12-01 00:00:00") - as.POSIXct("2015-11-01 00:00:00") # Time difference of 30.04167 days`

`r as.character(difftime(as.POSIXct("2015-12-01 00:00:00"), 
as.POSIXct("2015-11-01 00:00:00"), units = "sec")) # Time difference of 2595600 secs`

`r 2595600 - 2592000 == 3600`

`r 3600 == 60*60 # 1 hour`

## Accounting Date

> Definition: Accounting Date  
An accounting date is the cutoff date for 
reflecting events and recording amounts as paid or unpaid
in a financial statement or accounting system. 
The accounting date is sometimes referred to as the “as of” 
date.^[Casualty Actuarial Society *Statement
of Principles Regarding Property and Casualty Unpaid Claims Estimates*]

Although financial statements are always stated as of a date, 
say, December 31, 2015,
it would be more thorough to assign a time as well,
so as to identify all transactions
impacting the financial statements as of that date.
Some say "any particular moment on the 31st" could be used.^["it 
would be more accurate to write December 31, 20XX, 11:59:59, 
or any particular moment on the 
31st." http://www.investopedia.com/university/accounting/accounting5.asp]
However, payments can potentially occur up to and including midnight
December 31, 2015,
so if a time designator is to be used, 
midnight
would be the most appropriate.

Another reason for the closing instant to be the end of the day is
for the purposes of date arithmetic.
For example,
the instant closing the month of December is 31 days from the beginning
of December.

The International Organization for Standardization (ISO) sets
the standards for computer representation of
dates and times as a string of 
characters.^[ISO 8601:2004, 
*Data elements and interchange formats -- Information 
interchange -- Representation of dates and times*,
(c) ISO 2004]
Regarding "midnight",
the standards recognize that the instant 24:00:00 (hour 24)
marking the end of one calendar day 
coincides with the instant 00:00:00 (hour 00) marking the start of the next calendar day.
Furthermore, 
"the choice of [hour 24 or hour 00] will depend upon any association with a date,
or a time interval. 
[Hour 24] representations are preferred to represent the *end*
of a time interval."^[ISO 8601 4.2.3. Emphasis added.]
The concept of the closing of the books at the end of a calendar year, 
calendar month or calendar day^[defined as "time intervals" by ISO 8601]
suggests that hour 24 is preferred for representing
an accounting date.

ISO 8601's representation of the accounting date referred to as
"close of the 2015 calendar year" or "as of year end 2015"
is therefore "2015-12-31 24:00:00".

On the other hand,
base
R's
implementation of the 8601 standards^[The `POSIXt` classes. 
In R see ?DateTimeClasses. 
For additional background, see https://en.wikipedia.org/wiki/Unix_time]
stores time as the number of seconds since the beginning
of 1970,
so midnight is represented as hour 00 of the beginning of each 
day.
Therefore, the representation of "as of year end 2015"
in
R
is "2016-01-01 00:00:00"
which is January 1st when 
shortened to just the date.^[ISO 8601 2.1.5 recognizes that a "date" can represent 
an instant in time as well as a span of time ("duration").]
However,
to refer to the 
accounting date "as of year end 2015" with a label that
uses the month "January"
does not
satisfactorily communicate the
"as of December 31, 2015" accounting date concept.

It would be preferable that 
R
and a user be able to communicate that
concept using a label with "December" in it
and also have that label refer to the instant
in time separating calendar years 2015 and 2016 
(`as.Date("2016-01-01")` in
R).

That is the purpose of the "asof" class in the mondate package.

## Elapsed Month: "emonth"

> Definition: Age  
The age of event E as of accounting date A is the 
length of time from the occurrence of event E to 
hour 24 on accounting date A.

Note that units are not mentioned in the definition of age. 
This is because the appropriate unit will be specific to the use case.
In the vast majority of accounting cases,
age is measured in units of days, months or years, 
not seconds.

It turns out that the definitions of units days, months and years in ISO 8601
is not particularly straightforward,
with complexity in proportion to the width of the unit.
Due to the fundamental monthly, quarterly, and yearly accounting cycles,
it is generally not productive to used data defined more granularly than "month",
but even that level of granularity has its complexities.

This paper adds to that complexity with another take on "month": "emonth."



There is little ambiguity in defining elapsed time 
(durations) in units of
days.
ISO 8601 defines a 
R's
two Date/Time classes facilitate date/time
arithmetic in uits of seconds and days.

* Since R's POSIXt classes store time as the number of seconds
since the beginning of 1970,

so elapsed time in units of seconds is easily combined
with a POSIXt object to yield another POSIXt object unambiguously.

* The Date class stores time as the number of days
since the beginning of 1970,
so elapsed time in units of days is easily combined
with a Date object to yield another Date object unambiguously.

It is more complicated with units of months because
month periods are comprised of different numbers of days.
But this complication can be overcome with two key observations.

1. It is commonly accepted that the length of time between 
the beginning and end of a month is one "month".

2. Given any time *t* in any given month,
there is a unique portion *p*, 0<=*p*<=1,
that represents the portion of the month completed by time *t*.
Conversely, given any portion *p*,
the point in time of a month completed as of that portion of the month
can be easily found.

Those two observations give rise to the definition of the 
"month-time" of an instant in time:

>Definition: Month Time  
The month time of time *t* is a real number
measuring the number of months since the beginning of 1970 to time *t*.
The fractional part of the real number represents the portion
of *t*'s month to have expired by time *t*.



R
has two classes,
POSIXlt and POSIXct

### Sentence B

Base R follows that standard and represents an instant of time 
as the number of seconds that have transpired since the beginning of 1970.

It is well known that "months" are comprised of different 
numbers of days.
Although it is widely accepted that the length of time between 
the beginning and end of a month is one "month",
it is also recognized that month durations
cannot be defined consistently in units of days.
Those observations give rise to a different type of unit:
"elapsed month" or "emonth".

> Definition: Elapsed Month ("emonth")  
The length of time between two instants in time *t1* and *t2*
in units of "emonths" is  
  mondate(*t2*) - mondate(*t1*)
where
  *mondate(t)* is a real number representing
the number of months between the beginning of 1970 and
time *t*.

> Example  
mondate("1970-02-01") = 1  
mondate("1971-01-01") = 12
mondate("1970-02-15 00:00:00") = 1.5 because the beginning of 
February 15, 1970 is halfway through February 1970.
as.Date("1971-01-01") - as.Date("1971-02-01") = 10.5 emonths 

Conversely, given any real number t, 
it is straightforward to find the time ?YYYYMMDD?HH:MM:SS.zzzzz? 
that is t months away from the beginning of 1970 by 
first counting whole months, 
then counting into the next month (if necessary) the number of days and 
seconds corresponding to that month per the fractional value of t.

> Defintion: Elapsed Years  
The number of elapsed years between times *t1* and *t2*
equals the number of elapsed months divided by 12.

### Case Study:

ABC Ins. Co. started on 1/1/2010 to write earthquake insurance in California. 
As claims are made, 
ABC defines the occurrence date of a claim claim to be the date of 
the earthquake and stores that date in an ISO-8601 compliant o
bject (`POSIXt`). 
ABC has had good luck so far -- only 20 claims have been made. 
Here are their occurrence dates and their know values as of 12/31/2015:

data ?

To complete Sentence B, we need one more definition:

> Definition: The age of accident year AY as of AOD is AOD - AY-01-01.

The accident year age of the 20 claims is the vector  
	mondate(?2015-12-31?) ? year(occurrenceDate)-01-01



http://smallbusiness.chron.com/differences-dates-between-balance-sheet-income-sheet-24881.html

"Balance Sheet Date
A balance sheet often states that it is prepared as of a specific date, referred to as the balance sheet date. The balance sheet reports on a company’s financial conditions, namely the values of the company’s assets, liabilities and shareholders’ equity. Values are measured in terms of their monetary amounts at particular points in time rather than over any periods. At the end of an accounting cycle, with the accounting books closed to recording new business transactions, companies can summarize their financial conditions as of the cycle's end."

http://www.casact.org/professionalism/standards/princip/SOP-Regarding-Property-and-Casualty-Unpaid-Claims-Estimates_Final%204-22-2015.pdf

# Bibliography

Here's a pretty good paper about dates, Lubridate, Chron:
http://biostat.mc.vanderbilt.edu/wiki/pub/Main/ColeBeck/datestimes.pdf