---
title: "mondate Basics"
author: "Dan Murphy"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

In contrast to the two fundamental date/time classes that mark a day with the instant of
time at the beginning of the day,
a \code{mondate} represents a day as at the instant of time at the end of the day
("close of business").
This has various benefits, especially in the modeling of accounting entries' as-of dates.

The four major advantages of "mondate" are:

1. Date Aging
1. Flexible Date Formatting
1. Date Sequencing
1. Date Cutting

## Date Aging

The "age" of an event can be an important factor in a business use case.
On a personal note, 
my birth event took place on February 29, 1996. My age on February 28, 2015 was

```{r birthday}
require(mondate)
mondate("2015-02-28", timeunits = "years") - mondate("1996-02-29", timeunits = "years")
mondate("2015-02-28", timeunits = "months") - mondate("1996-02-29", timeunits = "months")
as.Date("2015-02-28") - as.Date("1996-02-29")
# or simply ..
YearsBetween("2015-02-28", "1996-02-29")

```

## Flexible Formatting

YearQuartersFormat

Note the various macros within the `vignette` section of the metadata block above. These are required in order to instruct R how to build the vignette. Note that you should change the `title` field and the `\VignetteIndexEntry` to match the title of your vignette.

## Styles

The `html_vignette` template includes a basic CSS theme. To override this theme you can specify your own CSS in the document metadata as follows:

    output: 
      rmarkdown::html_vignette:
        css: mystyles.css

## Figures

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
plot(1:10)
plot(10:1)
```

You can enable figure captions by `fig_caption: yes` in YAML:

    output:
      rmarkdown::html_vignette:
        fig_caption: yes

Then you can use the chunk option `fig.cap = "Your figure caption."` in **knitr**.

## More Examples

You can write math expressions, e.g. $Y = X\beta + \epsilon$, footnotes^[A footnote here.], and tables, e.g. using `knitr::kable()`.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(mtcars, 10))
```

Also a quote using `>`:

> "He who gives up [code] safety for [code] speed deserves neither."
([via](https://twitter.com/hadleywickham/status/504368538874703872))
